---
title: "Covid-19 Mental Health"
output: html_notebook
---

## Load packages
```{r}
library(zoo) # moving averages
library(plyr)
library(tidyverse)
library(readxl)
library(tibbletime)
library(lubridate)
```

Observational period between March 5 and September 5, 2020
```{r}
# volume data
cmh_ed <- read_excel('data/ED Volume Cambridge.xlsx')
grh_ed <- read_excel('data/ED volumes GRH.xlsx')
smh_ed <- read_excel('data/ED volumes SMH.xlsx')

# mental health data
cmh_mh <- read_excel('data/COVID-19 data sheet edited Jan 23.xlsx', sheet='cmh') %>%
  replace(is.na(.), 0) %>% 
  pivot_longer(cols = `March 1 - March 7`:`August 30th - September 5th`,
               names_to='week',
               values_to='count')

grh_mh <- read_excel('data/COVID-19 data sheet edited Jan 23.xlsx', sheet='grh') %>%
  replace(is.na(.), 0) %>% 
  pivot_longer(cols = `March 1 - March 7`:`August 30th - September 5th`,
               names_to='week',
               values_to='count')

smh_mh <- read_excel('data/COVID-19 data sheet edited Jan 23.xlsx', sheet='smh') %>%
  replace(is.na(.), 0) %>% 
  pivot_longer(cols = `March 1 - March 7`:`August 30th - September 5th`,
               names_to='week',
               values_to='count')

# Police data
pol <- read_excel('data/COVID-19 data sheet edited Jan 23.xlsx', sheet='pol') %>%
  filter(Subtype!='Drug overdose (monthly totals)' & Type!='EMS data') %>% 
  mutate(`June 21 - June 27`=as.double(`June 21 - June 27`)) %>% 
  replace(is.na(.), 0) %>%
  pivot_longer(cols = `March 1 - March 7`:`August 30th - September 5th`,
               names_to='week',
               values_to='count') 
```

## Figure 1
Get total volume
```{r}
vol <- smh_ed %>% left_join(grh_ed, by=c("Date of ED visit (reg)" = "Day of Triage_DT")) %>%
  left_join(cmh_ed, by=c("Date of ED visit (reg)" = "Triage Date")) %>% 
  rename(date=`Date of ED visit (reg)`, smh=`# of cases`, grh=`Number of Records`, cmh=Volume) %>% 
  mutate(total=smh+grh+cmh) %>% 
  mutate(year=as.factor(year(date)))
```

```{r}
vol %>% 
  group_by(year) %>% 
  summarize(sum(total))
```

```{r}
rbind(cmh_mh, grh_mh, smh_mh) %>% 
  group_by(Year) %>% 
  summarize(sum(count))
```

ED visits during lockdown  
Line graph
```{r}
vol <- vol %>% drop_na() %>% 
  group_by(year) %>% 
  mutate(roll = rollmean(total, k = 7, fill = NA)) %>% 
  mutate(date = if_else(year==2019, date+years(1), date))
p <- ggplot(vol, aes(x=date, y=total, color=year)) +
  geom_line() +
  geom_line(aes(x=date, y=roll)) +
  scale_color_manual(values=c("green", "blue")) +
  ylab('daily ED visits') +
  annotate('rect', xmin=as.POSIXct('2020-03-17'), xmax=as.POSIXct('2020-05-04'),
           ymin=-Inf, ymax=Inf, fill='red', alpha=0.3)
ggsave('graphs/vol.png', p, height = 8.5, width = 11, units = 'in')
p
```

```{r}
v2020 <- vol %>% 
  filter(year==2020)
max <- max(v2020$roll, na.rm = TRUE)
v2020$date[v2020$roll==max(v2020$roll, na.rm = TRUE)]
min <- min(v2020$roll, na.rm = TRUE)
v2020$date[v2020$roll==min(v2020$roll, na.rm = TRUE)]
(max-min)/max
```

Histogram (for potential ANOVA)
```{r}
vol %>% mutate(date=as.Date(date)) %>% 
  filter(date>='2019-03-17' & date<='2020-05-04') %>% 
  filter(date<='2019-05-04' | date>='2020-03-17') %>% 
  ggplot(aes(x=total, fill=year)) +
  geom_histogram()
```

Poisson ED visits during lockdown  
  
We're looking at the difference in outcome (hospitalizations) based on at least one predictor (year). That's general linear model territory. If you are summarizing over some period (e.g., the average across the shaded area), then the formulation of GLM that we refer to as ANOVA for shorthand would probably be OK. If you are not summarizing, and are using the individual data points, then you'll be violating some of the assumptions of GLM -- our observations are not independent, because the rate on Oct 2 is more related to the rate on Oct 3 than it is to April 3 (for example). In other words, these are time series data, where there is a trend happening, and to ignore that would violate the assumption of (residual) independence.  
  
Second, we might be violating the assumption that our residuals are normally distributed, because we know this type of data (counts) has been drawn from something other than a normal distribution, because only integers are possible and the counts are bounded at zero (normal distribution has limits of +/- infinity). So, we should be thinking about altering our GLM to expect a different distribution, like Poisson. However, it turns out that because the big problem with count data is the boundary at zero, when our counts are all far away from zero, the results of a model where we assume a normal distribution and where we assume a Poisson distribution converge. In the normal distribution, you get decimal points in your estimate, which don't really make sense (can't have half an ER visit), but this can easily be ignored. If you can live with that, then you might do the ANOVA. But, if you can stomach it, Poisson is the more correct version.  
  
In summary, can't use ANOVA because it violates two assumptions:  
1. The observations are not independent  
2. The residuals are not normally distributed  
  
We use Poisson regression because:  
1. It can handle observations that are not independent  
2. Counts are drawn from something other than a normal distribution, i.e. a Poisson distribution  
  
Before starting to interpret results, let’s check whether the model has over-dispersion or under-dispersion. If the Residual Deviance is greater than the degrees of freedom, then over-dispersion exists. This means that the estimates are correct, but the standard errors (standard deviation) are wrong and unaccounted for by the model.  
  
So, to have a more correct standard error we can use a quasi-poisson model:
```{r}
volume <- vol %>% mutate(date=as.Date(date)) %>% 
  filter(date>='2019-03-17' & date<='2020-05-04') %>% 
  filter(date<='2019-05-04' | date>='2020-03-17')

# Poisson regression
poisson_vol <- glm(total ~ year, family='quasipoisson', data=volume)
fig1 <- data.frame(coef(summary(poisson_vol))) %>% 
  slice(2) %>% 
  rename(`p-val(t)`=Pr...t..)
write_csv(fig1, 'stat_tests/fig1.csv')
summary(poisson_vol)
```

## Figure 2
Lockdown: 3/17 - 5/4
Weeks: 
```{r}
get_data <- function(data, type, subtype=NULL) {
  weeks <- c('March 15 - March 21\r\n\r\n', 'Mar 22 - Mar 28', 'Mar 29 - Apr 4', 'April 5 - 11', 'April 12 - 18', 'April 19 - 25', 'April 26 - May 2')
  if (is.null(subtype)) {
    smh <- data$smh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
    grh <- data$grh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
    cmh <- data$cmh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
  } else {
    smh <- data$smh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type, Subtype==subtype) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
    grh <- data$grh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type, Subtype==subtype) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
    cmh <- data$cmh_mh %>%
      filter(week %in% weeks) %>% 
      filter(Type==type, Subtype==subtype) %>% 
      group_by(Year, week) %>% 
      summarize(count=sum(count)) %>% 
      mutate(date=paste(week, Year, sep=' '))
  }
  whole <- smh %>% 
    left_join(grh, by='date') %>%
    left_join(cmh, by='date') %>% 
    mutate(total=count+count.x+count.y) %>% 
    select(Year, week, total) %>% 
    mutate(Year=as.factor(Year))
  return(whole)
}
```

Mental health data
```{r}
mh <- list(smh_mh=smh_mh, grh_mh=grh_mh, cmh_mh=cmh_mh)
```

Involuntary - Form 1 
```{r}
involuntary <- get_data(mh, 'Form 1 Total')

# Poisson regression
poisson_inv <- glm(total ~ Year, data=involuntary, family='poisson')
summary(poisson_inv)
```

Substance related
```{r}
substance <- get_data(mh, 'Substance related', 'Substance misuse/abuse')

# Poisson regression
poisson_sub <- glm(total ~ Year, data=substance, family='quasipoisson')
summary(poisson_sub)
```

Alcohol Related
```{r}
alcohol <- get_data(mh, 'Substance related', 'Alcohol (separate)')

# Poisson regression
poisson_alc <- glm(total ~ Year, data=alcohol, family='quasipoisson')
summary(poisson_alc)
```

Mood related
```{r}
mood <- get_data(mh, 'Mood related')

# Poisson regression
poisson_mood <- glm(total ~ Year, data=mood, family='quasipoisson')
summary(poisson_mood)
```

Psychosis related
```{r}
psychosis <- get_data(mh, 'Psychosis related')

# Poisson regression
poisson_psy <- glm(total ~ Year, data=psychosis, family='quasipoisson')
summary(poisson_psy)
```

Situational crisis
```{r}
situational <- get_data(mh, 'Situational crisis')

# Poisson regression
poisson_sit <- glm(total ~ Year, data=situational, family='poisson')
summary(poisson_sit)
```

Self harm related
```{r}
harm <- get_data(mh, 'Self harm')

# Poisson regression
poisson_harm <- glm(total ~ Year, data=harm, family='poisson')
summary(poisson_harm)
```

```{r}
get_fig2 <- function(poisson_fig2) {
  inv <- data.frame(coef(summary(poisson_fig2$poisson_inv))) %>% 
    slice(2) %>% 
    select(z.value, Pr...z..)
  sub <- data.frame(coef(summary(poisson_fig2$poisson_sub))) %>% 
    slice(2) %>% 
    select(t.value, Pr...t..)
  alc <- data.frame(coef(summary(poisson_fig2$poisson_alc))) %>% 
    slice(2) %>% 
    select(t.value, Pr...t..)
  mood <- data.frame(coef(summary(poisson_fig2$poisson_mood))) %>% 
    slice(2) %>% 
    select(t.value, Pr...t..)
  psy <- data.frame(coef(summary(poisson_fig2$poisson_psy))) %>% 
    slice(2) %>% 
    select(t.value, Pr...t..)
  sit <- data.frame(coef(summary(poisson_fig2$poisson_sit))) %>% 
    slice(2) %>% 
    select(z.value, Pr...z..)
  harm <- data.frame(coef(summary(poisson_fig2$poisson_harm))) %>% 
    slice(2) %>% 
    select(z.value, Pr...z..)
  
  tot <- rbind.fill(inv, sub, alc, mood, psy, sit, harm)
  tot$type <- c('Involuntary - Form 1', 'Substance related', 'Alcohol related', 'Mood related', 'Psychosis related', 'Situational crisis', 'Self harm related')
  tot <- tot %>% 
    rename(`p-val(z)`=Pr...z.., `p-val(t)`=Pr...t..)
  return(tot)
}
```

```{r}
poisson_fig2 <- list(poisson_inv=poisson_inv, poisson_sub=poisson_sub, poisson_alc=poisson_alc, poisson_mood=poisson_mood, poisson_psy=poisson_psy, poisson_sit=poisson_sit, poisson_harm=poisson_harm)
```

```{r}
fig2 <- get_fig2(poisson_fig2)
write_csv(fig2, 'stat_tests/fig2.csv')
```

## Figure 3

```{r}
weeks <- c('March 15 - March 21\r\n\r\n', 'Mar 22 - Mar 28', 'Mar 29 - Apr 4', 'April 5 - 11', 'April 12 - 18', 'April 19 - 25', 'April 26 - May 2')
```

Suicide
```{r}
suicide <- pol %>%
  filter(week %in% weeks) %>% 
  filter(Subtype=='Suicide attempt') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count))

# Poisson regression
poisson_sui <- glm(count ~ Year, data=suicide, family='poisson')
summary(poisson_sui)
```

Alcohol
```{r}
pol_alcohol <- pol %>%
  filter(week %in% weeks) %>% 
  filter(Subtype=='Intoxicated') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count))

# Poisson regression
poisson_pol_alc <- glm(count ~ Year, data=pol_alcohol, family='quasipoisson')
summary(poisson_pol_alc)
```

Assault
```{r}
assault <- pol %>%
  filter(week %in% weeks) %>% 
  filter(Subtype=='Assault') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count))

# Poisson regression
poisson_ass <- glm(count ~ Year, data=assault, family='quasipoisson')
summary(poisson_ass)
```

Domestic abuse
```{r}
domestic <- pol %>%
  filter(week %in% weeks) %>% 
  filter(Subtype=='Domestic dispute') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count))

# Poisson regression
poisson_dom <- glm(count ~ Year, data=domestic, family='quasipoisson')
summary(poisson_dom)
```

Crisis calls
```{r}
crisis <- pol %>%
  filter(week %in% weeks) %>% 
  filter(Subtype=='Crisis call volume') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count))

# Poisson regression
poisson_cri <- glm(count ~ Year, data=crisis, family='quasipoisson')
summary(poisson_cri)
```

```{r}

sui <- data.frame(coef(summary(poisson_sui))) %>% 
  slice(2) %>% 
  select(z.value, Pr...z..)
pol_alc <- data.frame(coef(summary(poisson_pol_alc))) %>% 
  slice(2) %>% 
  select(t.value, Pr...t..)
ass <- data.frame(coef(summary(poisson_ass))) %>% 
  slice(2) %>% 
  select(t.value, Pr...t..)
dom <- data.frame(coef(summary(poisson_dom))) %>% 
  slice(2) %>% 
  select(t.value, Pr...t..)
cri <- data.frame(coef(summary(poisson_cri))) %>% 
  slice(2) %>% 
  select(t.value, Pr...t..)

tot <- rbind.fill(sui, pol_alc, ass, dom, cri)
tot$type <- c('Suicide', 'Alcohol', 'Assault', 'Domestic Disputes', 'Crisis calls')
tot <- tot %>% 
  rename(`p-val(z)`=Pr...z.., `p-val(t)`=Pr...t..)

write_csv(tot, 'stat_tests/fig3.csv')
```

```{r}
week_order <- unique(pol$week)

crisis <- pol %>%
  filter(Subtype=='Crisis call volume') %>% 
  group_by(Year, week) %>% 
  summarize(count=sum(count)) %>% 
  mutate(Year=as.factor(Year)) %>% 
  mutate(week=factor(week, levels=week_order))
```

```{r}
p <- ggplot(crisis, aes(x=week, y=count, group=Year)) +
  geom_point(aes(color=Year)) +
  geom_line(aes(color=Year)) +
  theme(axis.text.x = element_text(angle =90, hjust = 1))

p
# library(segmented)
# data <- crisis %>% 
#   filter(Year==2020)
# my.lm <- lm(count ~ week, data = data)
# my.seg <- segmented(my.lm, seg.Z = ~ week)
# my.fitted <- fitted(my.seg)
# my.model <- data.frame(week = data$week, count = my.fitted) %>% 
#   mutate(week=factor(week, levels=week_order)) %>% 
#   mutate(Year=rep(2020, times=27))
# p + geom_line(my.model, mapping=aes(x=week, y=count))
```

```{r}
min <- crisis$count[crisis$Year==2020 & crisis$week=='May 31 - June 6']
max <- crisis$count[crisis$Year==2020 & crisis$week=='July 12 - July 18']
(max-min)/min
```

```{r}
inv <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Involuntary Form 1'))
sub <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Substance related'))
alc <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Alcohol related'))
mood <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Mood related'))
psy <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Psychosis related'))
sit <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Situational crisis'))
self <- glm.nb(Count ~ Year, data=mh_lock %>% filter(Type=='Self harm related'))

summary(inv)
summary(sub)
summary(alc)
summary(mood)
summary(psy)
summary(sit)
summary(self)
```

```{r}
sui <- glm.nb(Count ~ Year, data=pol_lock %>% filter(Type=='Suicide attempt'))
int <- glm.nb(Count ~ Year, data=pol_lock %>% filter(Type=='Intoxication'))
ass <- glm.nb(Count ~ Year, data=pol_lock %>% filter(Type=='Assault'))
dom <- glm.nb(Count ~ Year, data=pol_lock %>% filter(Type=='Domestic dispute'))
cri <- glm.nb(Count ~ Year, data=pol_lock %>% filter(Type=='Crisis calls'))

summary(sui)
summary(int)
summary(ass)
summary(dom)
summary(cri)
```

```{r}
mh_poisson <- glm(Count ~ Year * Type, data = mh_lock, family = 'poisson')
mh_nb <- glm.nb(Count ~ Year * Type, data = mh_lock)

summary(mh_poisson)
summary(mh_nb)
```

```{r}
pol_poisson <- glm(Count ~ Year + Type, data = pol_lock, family = 'poisson')
pol_nb <- glm.nb(Count ~ Year + Type, data = pol_lock)

summary(pol_poisson)
summary(pol_nb)
```

```{r}
out=cpt.mean(x, pen.value = c(2*log(length(x)),100*log(length(x))),penalty = "CROPS",method = "PELT") 
cpts.full(out)  # returns 7 segmentations for penalty values between 4 and 1500.  
# We find segmentations with 7, 5, 4, 3, 2, 1 and 0 changepoints. 
# Note that the empty final row indicates no changepoints.
pen.value.full(out) # gives associated penalty transition points
# CROPS does not give an optimal set of changepoints thus we may wish to explore further
plot(out,diagnostic=TRUE) 
# looks like the segmentation with 3 changepoints, 50,100,150 is the most appropriate
plot(out,ncpts=12)
```


```{r}
y_ts    <- ts(x)

cptfn <- function(data, pen) {
  ans <- cpt.mean(data, test.stat="Normal", method = "PELT", penalty = "Manual", pen.value = pen) 
  length(cpts(ans)) +1
}
 
# evaluate and plot results:
plot.new()
frame()
# run cptfn for the signal with a known change point
pen.vals <- seq(0, 10000, 100)
elbowplotData <- unlist(lapply(pen.vals, function(p) 
                  cptfn(data = y_ts, pen = p)))
plot.ts(y_ts,type='l',col='red',
     xlab = "time",
     ylab = " Y(t)",
     main = "Stationary signal (constant mean)")
plot(pen.vals,elbowplotData, 
     xlab = "PELT penalty parameter",
     ylab = " ",
     main = " ")
```

```{r}
penalty.val <- 50000 # this value is determined from elbow plots

cptm_stationary <- cpt.mean(y_ts,    penalty='Manual',pen.value=penalty.val,method='PELT') 
cpts_stationary <- cpts(cptm_stationary) # change point time points

cpts_stationary
```

```{r}
plot.new()
frame()
par(mfcol=c(1,2))
plot(cptm_stationary,
     xlab = "time",
     ylab = " Y(t)",
     main = "Change in mean signal")
```


```{r}
library(segmented)
my.lm <- lm(total ~ date, data = vol)
# make segmented regression
my.seg <- segmented(my.lm, seg.Z = ~ date)
my.fitted <- fitted(my.seg)

vol %>% 
  mutate(fitted = my.fitted) %>% 
  ggplot() +
  geom_line(mapping = aes(x=date, y=total)) +
  geom_line(mapping = aes(x=date, y=fitted), color='blue')
```


